---
title: "FP4 Ethan.RMD"
output: html_document
date: "2024-11-14"
author: "Ethan Caldecott"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# *Focus*: My job is to look at frequency of use look as my quality of life outcome after treatment. Is there improvement in frequency of use between time of admission and disharge?

# Libraries
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(haven)
library(scales)
library(ggalluvial)
library(stringr)
library(randomForest)
library(ggmosaic)
library(viridis)
library(tidymodels) 
library(ranger) 
library(vip)
library(maps)
```

# Load Data
## Link to Codebook: https://www.samhsa.gov/data/system/files/media-puf-file/TEDS-D-2021-DS0001-info-codebook.pdf
```{r}
load("C:/Users/ethan/OneDrive/Desktop/STAT 456/TEDS-D-2021 October 28/tedsd_puf_2021_r.rdata")
```

# Randomly Exploring the Data
```{r}
summary(tedsd_puf_2021_r)
```

# Intermediate Presentation FP4
```{r}
# Reshape data for comparison
tedsd_long <- tedsd_puf_2021_r %>%
  select(FREQ1, FREQ1_D, FREQ2, FREQ2_D, FREQ3, FREQ3_D) %>%
  mutate(across(everything(), haven::as_factor))

# Plot stacked bar chart
ggplot(tedsd_long, aes(x = FREQ1, fill = FREQ1_D)) +
  geom_bar(position = "fill") +
  labs(title = "Frequency of Use at Admission vs. Discharge",
       y = "Proportion", x = "Admission",
       fill = "Frequency Level at Discharge") +
  theme_minimal() 

# Plot stacked bar chart
ggplot(tedsd_long, aes(x = FREQ2, fill = FREQ2_D)) +
  geom_bar(position = "fill") +
  labs(title = "Frequency of Use at Admission vs. Discharge",
       y = "Proportion", x = "Admission",
       fill = "Frequency Level at Discharge") +
  theme_minimal() 

# Plot stacked bar chart
ggplot(tedsd_long, aes(x = FREQ3, fill = FREQ3_D)) +
  geom_bar(position = "fill") +
  labs(title = "Frequency of Use at Admission vs. Discharge",
       y = "Proportion", x = "Admission",
       fill = "Frequency Level at Discharge") +
  theme_minimal() 
```
## Data Filters
```{r}
freq1_data <- tedsd_puf_2021_r %>%
  select(FREQ1, FREQ1_D, SUB1, SERVICES_D, LOS, REASON) %>%
  mutate(across(everything(), as.factor)) %>%
  rename(Freq = FREQ1, Freq_d = FREQ1_D, Sub = SUB1, Service_D = SERVICES_D) %>%
  mutate(Freq_Type = "Primary") %>%
  filter(Service_D != "1" | Service_D != "2") %>% 
  filter(REASON == "1") %>% 
  filter(Freq != -9 & Freq_d != -9 & Sub != -9) %>%
  mutate(LOS = as.integer(LOS)) %>%
  mutate(
    Freq_Label = case_when(
      Freq %in% c("1") ~ "No use in the past month",
      Freq %in% c("2") ~ "Some use",
      Freq %in% c("3") ~ "Daily use"  
    ),
    Freq_d_Label = case_when(
      Freq_d %in% c("1") ~ "No use in the past month",
      Freq_d %in% c("2") ~ "Some use",
      Freq_d %in% c("3") ~ "Daily use"  
    ),
    Service_D_Label = case_when(
      Service_D %in% c("1") ~ "Detox, 24-hour, hospital inpatient",
      Service_D %in% c("2")  ~ "Detox, 24-hour, free-standing residential",
      Service_D %in% c("3")  ~ "Rehab/residential, hospital (non-detox)",
      Service_D %in% c("4")  ~ "Rehab/residential, short term (30 days or fewer)",
      Service_D %in% c("5")  ~ "Rehab/residential, long term (more than 30 days)",
      Service_D %in% c("6")  ~ "Ambulatory, intensive outpatient",
      Service_D %in% c("7")  ~ "Ambulatory, non-intensive outpatient",
      Service_D %in% c("8")  ~ "Ambulatory, detoxification",
      TRUE ~ "Other"),
    LOS_Label = case_when(
      LOS == 31 ~ "31 to 45 days",
      LOS == 32 ~ "46 to 60 days",
      LOS == 33 ~ "61 to 90 days",
      LOS == 34 ~ "91 to 120 days",
      LOS == 35 ~ "121 to 180 days",
      LOS == 36 ~ "181 to 365 days",
      LOS == 37 ~ "More than a year",
      TRUE ~ "Other"
    )) %>% # Sorts Labels for Graphs
  mutate(
    Freq_d_Label = factor(Freq_d_Label, levels = c("No use in the past month", "Some use", "Daily use")),
    Freq_Label = factor(Freq_Label, levels = c("No use in the past month", "Some use", "Daily use")))  %>%
  filter(LOS_Label != "Other")
  
```

## All Substance Data
```{r}
freq1_data_all <- freq1_data %>%
    mutate(Sub_Label = "All Substances")  
```

## Alcohol, Opioids, Stimulants (AOS) Data 
```{r}
freq1_data_top3 <- freq1_data %>%
  mutate(Sub_Label = case_when(
      Sub %in% c("2") ~ "Alcohol",
      Sub %in% c("5", "7") ~ "Opioids",
      Sub %in% c("3", "10", "12") ~ "Stimulants",
      TRUE ~ "Other")) %>% 
  filter(Sub_Label != "Other") 
```

## Combine All Substance & AOS Data
```{r}
freq1_data_allANDtop3 <- rbind(freq1_data_all, freq1_data_top3) %>%
  mutate(Sub_Label = factor(Sub_Label, levels = c("All Substances", "Alcohol", "Opioids", "Stimulants")))
```

## Graph 1: Change in Frequency of Use from Admission to Discharge
```{r}
# Data Code
graph1part1 <- freq1_data_allANDtop3 %>%
  group_by(Sub_Label, Freq_d_Label) %>%
  count(name = "Freq_d_Count") %>%         
  ungroup() %>%
  group_by(Sub_Label) %>%
  mutate(Total_Count = sum(Freq_d_Count)) %>%
  mutate(Percentage_d = Freq_d_Count/Total_Count*100)

graph1part2 <- freq1_data_allANDtop3 %>%
  group_by(Sub_Label, Freq_Label) %>% 
  count(name = "Freq_Count") %>% # Count each Frequency and Substance pairing
  ungroup() %>%
  group_by(Sub_Label) %>% # Group by Substance
  mutate(
    Total_Count = sum(Freq_Count), # Total count for each Substance
    Percentage = (Freq_Count / Total_Count) * 100 # % for each Frequency 
  )

percentage_diff <- graph1part2 %>%
  inner_join(graph1part1, by = c("Sub_Label", "Freq_Label" = "Freq_d_Label")) %>%
  mutate(
    Percent_dminusa = Percentage_d - Percentage
  ) %>%
  select(Sub_Label, Freq_Label, Percent_dminusa, Percentage_d, Percentage)

# Graph Code
ggplot(percentage_diff, aes(x = Freq_Label, y = Percent_dminusa)) +
  geom_col(aes(fill = Freq_Label)) + 
  labs(
    title = "Change in Frequency of Use from Admission to Discharge",
    subtitle = "The % of use at Discharge - % of use at Addmission",
    x = "",
    y = "Difference in Percentage (%)",
    fill = "Frequency Level"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") + 
  geom_text(aes(label = paste0(round(Percent_dminusa), "%")),  position = position_stack(vjust = 0.5), size = 5) +
  scale_fill_manual(values = c("No use in the past month" = "green", "Some use" = "yellow", "Daily use" = "red")) +
  facet_wrap(~ Sub_Label,ncol = 1) +  
  theme_minimal() +
  coord_flip()+
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme(legend.position = "top",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16),
    plot.title = element_text(face = "bold", size = 24),     
    plot.subtitle = element_text(size = 18),
    strip.text = element_text(size = 16, face = "bold"))

```

## Graph 2: Comparison of Frequency Levels at Admission and Discharge
```{r}
# Data Code
stackedbar_data <- percentage_diff %>%
  gather(key = "Stage", value = "Percentage", Percentage, Percentage_d) %>%
  mutate(Stage = ifelse(Stage == "Percentage", "Admission", "Discharge"))

# Graph Code
ggplot(stackedbar_data, aes(x = Stage, y = Percentage, fill = Freq_Label)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage, 0), "%")), 
            position = position_stack(vjust = 0.5), size = 5, color = "black") +
  facet_wrap(~ Sub_Label) +
  labs(
    title = "Comparison of Frequency Levels at Admission and Discharge",
    y = "Percentage (%)",
    fill = "Frequency Level",
    x = " "
  ) +
  scale_fill_manual(values = c("No use in the past month" = "green", "Some use" = "yellow", "Daily use" = "red")) +
  theme_minimal() +
  theme(legend.position = "top",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16),
    plot.title = element_text(face = "bold", size = 24),     
    plot.subtitle = element_text(size = 18),
    strip.text = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 16, face = "bold"))

```

## Graph 3: *Ambulatory* Comparison of Frequency Levels at Admission and Discharge
```{r}
# Data Code
graph1part1 <- freq1_data_allANDtop3 %>% 
  filter(str_starts(Service_D_Label, "Ambulatory")) %>% # Ambulatory ONLY
  group_by(Sub_Label, Freq_d_Label) %>%
  count(name = "Freq_d_Count") %>%         
  ungroup() %>%
  group_by(Sub_Label) %>%
  mutate(Total_Count = sum(Freq_d_Count)) %>%
  mutate(Percentage_d = Freq_d_Count/Total_Count*100)

graph1part2 <- freq1_data_allANDtop3 %>%
  filter(str_starts(Service_D_Label, "Ambulatory")) %>%
  group_by(Sub_Label, Freq_Label) %>%                     
  count(name = "Freq_Count") %>%                    
  ungroup() %>%
  group_by(Sub_Label) %>%                          
  mutate(
    Total_Count = sum(Freq_Count),                  
    Percentage = (Freq_Count / Total_Count) * 100  
  )

percentage_diff <- graph1part2 %>%
  inner_join(graph1part1, by = c("Sub_Label", "Freq_Label" = "Freq_d_Label")) %>%
  mutate(
    Percent_dminusa = Percentage_d - Percentage
  ) %>%
  select(Sub_Label, Freq_Label, Percent_dminusa, Percentage_d, Percentage)

stackedbar_data <- percentage_diff %>%
  gather(key = "Stage", value = "Percentage", Percentage, Percentage_d) %>%
  mutate(Stage = ifelse(Stage == "Percentage", "Admission", "Discharge"))

# Graph Code
ggplot(stackedbar_data, aes(x = Stage, y = Percentage, fill = Freq_Label)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage, 0), "%")), 
            position = position_stack(vjust = 0.5), size = 5, color = "black") +
  facet_wrap(~ Sub_Label) +
  labs(
    title = "*Residential Rehab* Comparison of Frequency Levels at Admission and Discharge",
    y = "Percentage (%)",
    fill = "Frequency Level",
    x = " "
  ) +
  scale_fill_manual(values = c("No use in the past month" = "green", "Some use" = "yellow", "Daily use" = "red")) +
  theme_minimal() +
  theme(legend.position = "top",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16),
    plot.title = element_text(face = "bold", size = 20),     
    plot.subtitle = element_text(size = 18),
    strip.text = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 16, face = "bold"))
```
## Graph 4: *Residential Rehab* Comparison of Frequency Levels at Admission and Discharge
```{r}
# Data Code
graph1part1 <- freq1_data_allANDtop3 %>%
  filter(!str_starts(Service_D_Label, "Ambulatory")) %>% # Non-Ambulatory aka Residential Rehab ONLY
  group_by(Sub_Label, Freq_d_Label) %>%
  count(name = "Freq_d_Count") %>%         
  ungroup() %>%
  group_by(Sub_Label) %>%
  mutate(Total_Count = sum(Freq_d_Count)) %>%
  mutate(Percentage_d = Freq_d_Count/Total_Count*100)

graph1part2 <- freq1_data_allANDtop3 %>%
  filter(!str_starts(Service_D_Label, "Ambulatory")) %>%
  group_by(Sub_Label, Freq_Label) %>% 
  count(name = "Freq_Count") %>% 
  ungroup() %>%
  group_by(Sub_Label) %>% 
  mutate(
    Total_Count = sum(Freq_Count),                  
    Percentage = (Freq_Count / Total_Count) * 100   
  )

percentage_diff <- graph1part2 %>%
  inner_join(graph1part1, by = c("Sub_Label", "Freq_Label" = "Freq_d_Label")) %>%
  mutate(
    Percent_dminusa = Percentage_d - Percentage
  ) %>%
  select(Sub_Label, Freq_Label, Percent_dminusa, Percentage_d, Percentage)

stackedbar_data <- percentage_diff %>%
  gather(key = "Stage", value = "Percentage", Percentage, Percentage_d) %>%
  mutate(Stage = ifelse(Stage == "Percentage", "Admission", "Discharge"))

# Graph Code
ggplot(stackedbar_data, aes(x = Stage, y = Percentage, fill = Freq_Label)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage, 0), "%")), 
            position = position_stack(vjust = 0.5), size = 5, color = "black") +
  facet_wrap(~ Sub_Label) +
  labs(
    title = "*Residential Rehab* Comparison of Frequency Levels at Admission and Discharge",
    y = "Percentage (%)",
    fill = "Frequency Level",
    x = " "
  ) +
  scale_fill_manual(values = c("No use in the past month" = "green", "Some use" = "yellow", "Daily use" = "red")) +
  theme_minimal() +
  theme(legend.position = "top",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16),
    plot.title = element_text(face = "bold", size = 20),     
    plot.subtitle = element_text(size = 18),
    strip.text = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 16, face = "bold"))
```

# Data for Final
## Labels included (Dataset Name: Description)
AGE: Age at admission
GENDER: Gender
RACE: Race
ETHNIC: Ethnicity
MARSTAT: Marital status
EDUC: Education at admission
EMPLOY: Employment status at admission
FREQ1: Frequency of use at admission (primary)
FREQ1_D: Frequency of use at discharge (primary)
SUB1: Primary substance at admission
```{r}
data1 <- tedsd_puf_2021_r %>% 
  select(-c(DISYR, CASEID, ALCFLG, COKEFLG, MARFLG, HERFLG, METHFLG, OPSYNFLG, PCPFLG, HALLFLG, MTHAMFLG, AMPHFLG, STIMFLG, BENZFLG, TRNQFLG, BARBFLG, SEDHPFLG, INHFLG, OTCFLG, OTHERFLG, ALCDRUG, DETNLF, DETNLF_D, PREG, PRIMINC, CBSA2020, DAYWAIT, DETCRIM, ROUTE2, FREQ2, FREQ2_D, FRSTUSE2, ROUTE3, FREQ3, FREQ3_D, FRSTUSE3, PRIMINC, PRIMPAY)) %>%
  filter(FREQ1_D != -9) %>%
  filter(FREQ1 != -9) %>%
  #  filter(REASON == 1) %>%
  select(
    # Target variable (1)
    FREQ1_D,
    # Demographic Variables (7)
    AGE, GENDER, RACE, ETHNIC, EDUC, MARSTAT, VET,
    # Other Outcome Variabl+s (5)
    EMPLOY, EMPLOY_D, LIVARAG, LIVARAG_D, FREQ1,
    # Non-Treatment Variables (8)
    # ROUTE1, DUE TO REPETIVENES 
    FRSTUSE1, DSMCRIT, PSYPROB, PSOURCE, STFIPS, ARRESTS_D,
    HLTHINS, NOPRIOR, 
    # Treatment-Related Variables (7)
    SUB1, 
    SUB2, SUB3, # BINARY YES/NO FILTER 1 NO
    METHUSE, LOS, SERVICES, FREQ_ATND_SELF_HELP_D
    ) %>%
  mutate(across(everything(), haven::as_factor)) %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(across(where(is.numeric), ~ replace_na(., 0)))
data1$FREQ1_D <- droplevels(data1$FREQ1_D)
data1[data1 == -9] <- NA
```

## Check stuff
```{r}
colnames(data1)
table(data1$FREQ1_D)
table(data1$FREQ1)

```
## Random Forest
```{r}
rf_spec <- rand_forest() %>%
  set_engine(engine = 'ranger') %>% 
  set_args(mtry = NULL, 
           trees = 500, 
           min_n = 2,
           probability = FALSE, 
           importance = 'impurity') %>% 
  set_mode('classification') 

data_rec <- recipe(FREQ1_D ~ ., data = data1) # predict employment at discharge

data_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(data_rec)

rf_fit <- fit(data_wf, data = data1)

rf_fit
```

### Random Forest Variable Importance
```{r}
#Variable importance
rf_fit %>% 
    extract_fit_engine() %>% 
    vip(num_features = 6, mapping = aes(fill = .data[["Variable"]])) +
    scale_fill_manual(values = c("grey", "grey", "grey" ,"#A94064" ,"#A94064" ,"grey" )) +
    labs(title = "Top 6 Most Important Variables ",
         subtitle = "in Predicting Frequency of Use at Discharge") +
    theme_classic() +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", size = 11),
          axis.text.y = element_blank(),       # Remove y-axis text
          axis.ticks.y = element_blank(),      # Remove y-axis ticks
          axis.line.y = element_blank())+    # Remove y-axis line)+
    geom_text(aes(label = c("FREQUENCY OF USE AT ADMISSION",
                            "STATE",
                            "LENGTH OF STAY",
                            "SERVICES AT ADMISSION",
                            "AGE",
                            "AGE AT FIRST USE"), 
                            y = 0), hjust = 0, color = "black", size = 4, fontface = "bold")
```

### Accuracy Score of RF
```{r}
#accuracy of random forest model
rf_OOB_output <- function(fit_model, model_label, truth){
    tibble(
          .pred_Unemployed = fit_model %>% extract_fit_engine() %>% pluck('predictions'), #OOB predictions
          FREQ1_D= truth,
          label = model_label
      )
}

output <- rf_OOB_output(rf_fit, "test", data1  %>% pull(FREQ1_D))

output %>% accuracy(truth = FREQ1_D, estimate = .pred_Unemployed) # print accuracy
```



##  Mosaic LOS
```{r}
los_mosaic <- data1 %>%
  mutate(
    FREQ1 = as.numeric(factor(FREQ1, 
                              levels = c("No use in the past month", 
                                         "Some use", 
                                         "Daily use"))),
    FREQ1_D = as.numeric(factor(FREQ1_D, 
                                levels = c("No use in the past month", 
                                           "Some use", 
                                           "Daily use"))),
    EMPLOYMENT_CHANGE = case_when(
      FREQ1 == FREQ1_D ~ "Same",
      FREQ1 < FREQ1_D ~ "Worse",
      FREQ1 > FREQ1_D ~ "Improved"
    ),
    lengthOfStay = case_when(
      LOS %in% c("91 to 120 days", "121 to 180 days", "181 to 365 days", "More than a year") ~ "> 90 days",
      LOS %in% c("30", "31 to 45 days", "46 to 60 days", "61 to 90 days") ~ "30-90 days",
      TRUE ~ "< 30 days"  # Default case for all other values
    )
  )%>%
  mutate(lengthOfStay = factor(lengthOfStay, levels = c("< 30 days" , "30-90 days", "> 90 days"))) 


los_mosaic %>%
  ggplot() +
  geom_mosaic(aes(x = product(EMPLOYMENT_CHANGE, lengthOfStay), fill = EMPLOYMENT_CHANGE)) +
  geom_mosaic_text(aes(
    x = product(EMPLOYMENT_CHANGE, lengthOfStay), 
    label = paste0(round(after_stat(.wt) / sum(after_stat(.wt)) * 100, 1), "%")
  )) +
  labs(
    fill = "Usage Change",
    y = "Proportion of Substance Use Patients",
    x = "Length of Stay",
    title = "Impact of Length of Stay on Employment Change",
    subtitle = "Among Substance Use Patients in Public Facilities (2021)"
  ) +
  scale_fill_manual(
    values = c(
      "Improved" = "darkgreen",  # Green for improvement
      "Same" = "#D3D3D3",        # Gray for same
      "Worse" = "red"            # Red for worse
    )
  ) +
  theme_classic() +
  coord_flip() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11)
  )
```

##  Mosaic Service
```{r}

service_mosaic <- data1 %>%
  mutate(
    FREQ1 = as.numeric(factor(FREQ1, 
                              levels = c("No use in the past month", 
                                         "Some use", 
                                         "Daily use"))),
    FREQ1_D = as.numeric(factor(FREQ1_D, 
                                levels = c("No use in the past month", 
                                           "Some use", 
                                           "Daily use"))),
    EMPLOYMENT_CHANGE = case_when(
      FREQ1 == FREQ1_D ~ "Same",
      FREQ1 < FREQ1_D ~ "Worse",
      FREQ1 > FREQ1_D ~ "Improved"
    ),
    lengthOfStay = case_when(
      SERVICES %in% c(
        "Detox, 24-hour, hospital inpatient",
        "Detox, 24-hour, free-standing residential",
        "Ambulatory, detoxification"
      ) ~ "Detox",
      
      SERVICES %in% c(
        "Rehab/residential, hospital (non-detox)",
        "Rehab/residential, short term (30 days or fewer)",
        "Rehab/residential, long term (more than 30 days)"
      ) ~ "Rehab/Residential",
      
      SERVICES %in% c(
        "Ambulatory, intensive outpatient",
        "Ambulatory, non-intensive outpatient"
      ) ~ "Ambulatory"))

service_mosaic %>%
  ggplot() +
  geom_mosaic(aes(x = product(lengthOfStay, EMPLOYMENT_CHANGE), fill = lengthOfStay)) +
  geom_mosaic_text(aes(x = product(lengthOfStay, EMPLOYMENT_CHANGE), label = scales::percent(after_stat(.wt)/ sum(after_stat(.wt))))) +
   labs(fill = "Length of Treatment", y = "Proportion of Substance Use Patients", x = "Frequency of Use", title = "Impact of Length of Stay on Frequency of Use", subtitle = "Among Substance Use Patients in Public Facilities (2021)") +
  # scale_fill_manual(values = c("#2A788E", "#FDE725")) +
  scale_fill_viridis_d() +
  theme_classic() +
  coord_flip() +
  theme(legend.position = "bottom", axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11)) 

```

## Final 1: Proportions 3 Level LOS
```{r}
# Compute proportions
plot_data <- los_mosaic %>%
  count(lengthOfStay, EMPLOYMENT_CHANGE) %>%
  group_by(lengthOfStay) %>%
  mutate(proportion = n / sum(n)) %>%
  mutate(EMPLOYMENT_CHANGE = factor(EMPLOYMENT_CHANGE, levels = c("Worse", "Same", "Improved")))


ggplot(plot_data, aes(x = lengthOfStay, y = proportion, fill = EMPLOYMENT_CHANGE)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = scales::percent(proportion, accuracy = 0.1)), 
            position = position_fill(vjust = 0.5), size = 5, fontface = "bold") +
  labs(title = "Impact of Length of Stay on Frequency of Use", 
       subtitle = "Among Substance Use Patients in Public Facilities (2021)",
       x = "Length of Stay",
       y = "Percentage of Patients",
       fill = "Frequency of Use Change") +
  scale_fill_manual(values = c("Improved" = "darkgreen", "Same" = "gray", "Worse" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11),
        axis.title.x = element_text(size = 14, face = "bold"), 
        axis.text.x = element_text(size = 12, face = "bold"))


```


## Final 2: Proportions 3 Level Service
```{r}
# Compute proportions
plot_data <- service_mosaic %>%
  count(lengthOfStay, EMPLOYMENT_CHANGE) %>%
  group_by(lengthOfStay) %>%
  mutate(proportion = n / sum(n)) %>%
  mutate(EMPLOYMENT_CHANGE = factor(EMPLOYMENT_CHANGE, levels = c("Worse", "Same", "Improved")))

# Create the plot
ggplot(plot_data, aes(x = lengthOfStay, y = proportion, fill = EMPLOYMENT_CHANGE)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = scales::percent(proportion, accuracy = 0.1)), 
            position = position_fill(vjust = 0.5), size = 5, fontface = "bold") +
  labs(title = "Impact of Treatment Service on Frequency of Use", 
       subtitle = "Among Substance Use Patients in Public Facilities (2021)",
       x = "Treatment Service",
       y = "Percentage of Patients",
       fill = "Frequency of Use Change") +
  scale_fill_manual(values = c("Improved" = "darkgreen", "Same" = "gray", "Worse" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11),
        axis.title.x = element_text(size = 14, face = "bold"), 
        axis.text.x = element_text(size = 12, face = "bold"))


```
## State Map Avg Usage Worse to Improve
```{r}
los_map <- los_mosaic %>%
  rename(state = STFIPS) %>% 
  mutate(EMPLOYMENT_CHANGE_counter = case_when(
         EMPLOYMENT_CHANGE == "Same" ~ 0,   
         EMPLOYMENT_CHANGE == "Worse" ~ -1, 
         EMPLOYMENT_CHANGE == "Improved" ~ 1, 
         TRUE ~ NA_real_)) %>%
  group_by(state) %>%                
  summarize(value = mean(EMPLOYMENT_CHANGE_counter, na.rm = TRUE)) %>% 
  mutate(state = tolower(state))    

us_map <- map_data("state")

map_data <- us_map %>%
  left_join(los_map, by = c("region" = "state"))


state_centers <- map_data %>%
  group_by(region) %>%
  summarize(long = mean(range(long, na.rm = TRUE)),
            lat = mean(range(lat, na.rm = TRUE)),
            value = mean(value, na.rm = TRUE)) %>%
  ungroup() 

ggplot(map_data, aes(x = long, y = lat, group = group, fill = value)) +
  geom_polygon(color = "black") + 
  scale_fill_gradient2(low = "red",    
                       mid = "white",     
                       high = "darkgreen",     
                       midpoint = 0,      
                       na.value = "gray90") +
  geom_text(data = state_centers,
            aes(x = long, y = lat, label = round(value, 2)), inherit.aes = FALSE, color = "black", size = 3) +
  labs(title = "Choropleth Map of Employment Changes by State",
       fill = "Average Change",
       caption = "Values range from -1 (Worse) to 1 (Improved)", # Add caption
       x = "",
       y = "") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())




```

## Proportion All Levels of Services
```{r}
los_mosaic %>%
  filter(FREQ1 != "1") %>%
  count(SERVICES, EMPLOYMENT_CHANGE) %>%
  group_by(SERVICES) %>%
  mutate(proportion = n / sum(n)) %>%
  mutate(EMPLOYMENT_CHANGE = factor(EMPLOYMENT_CHANGE, levels = c("Worse", "Same", "Improved"))) %>%
  ggplot(aes(x = SERVICES, y = proportion, fill = EMPLOYMENT_CHANGE)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Proportion of Frequency of Use Outcomes by Service Type",
       x = "Service Type",
       y = "Proportion",
    fill = "Frequency of Use Outcome") +
  scale_fill_manual(values = c("Improved" = "darkgreen", "Same" = "#D3D3D3", "Worse" = "red")) +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Mosaic All Levels of Service
```{r}
combined_mosaic <- los_mosaic %>%
  mutate(SERVICES3 = case_when(
      SERVICES %in% c("Detox, 24-hour, hospital inpatient",
                      "Detox, 24-hour, free-standing residential",
                      "Ambulatory, detoxification") ~ "Detox",
      SERVICES %in% c("Rehab/residential, hospital (non-detox)",
                      "Rehab/residential, short term (30 days or fewer)",
                      "Rehab/residential, long term (more than 30 days)") ~ "Rehab/Residential",
      SERVICES %in% c("Ambulatory, intensive outpatient",
                      "Ambulatory, non-intensive outpatient") ~ "Ambulatory")) %>%
  mutate(SERVcommaLOS = paste(SERVICES3, lengthOfStay,sep = ", ")) %>% 
  mutate(SERVcommaLOS = factor(SERVcommaLOS,
      levels =c("Ambulatory, < 30 days",
                "Ambulatory, 30-90 days",
                "Ambulatory, > 90 days",
                "Detox, < 30 days",
                "Detox, 30-90 days",
                "Detox, > 90 days",
                "Rehab/Residential, < 30 days",
                "Rehab/Residential, 30-90 days",
                "Rehab/Residential, > 90 days")))

combined_mosaic %>%
  ggplot() +
  geom_mosaic(aes(x = product(EMPLOYMENT_CHANGE, SERVcommaLOS), fill = EMPLOYMENT_CHANGE)) +
  geom_mosaic_text(aes(x = product(EMPLOYMENT_CHANGE, SERVcommaLOS), label = paste0(round(after_stat(.wt) / sum(after_stat(.wt)) * 100, 1), "%"))) +
  labs(fill = "Usage Change",
       y = "Proportion of Substance Use Patients",
       x = "Length of Stay",
       title = "Impact of Length of Stay on Employment Change",
       subtitle = "Among Substance Use Patients in Public Facilities (2021)") +
  scale_fill_manual(values = c("Improved" = "darkgreen", "Same" = "#D3D3D3", "Worse" = "red"))+
    theme_classic() +
  coord_flip() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11))
```
## Final: EMPLOY_D Distribution
```{r}
data1 %>%
  filter(EMPLOY_D != -9) %>%
  mutate(EMPLOY_D = ifelse(EMPLOY_D %in% c("Part-time", "Full-time"), "Employed", EMPLOY_D)) %>%
  mutate(EMPLOY_D = case_when(EMPLOY_D == 5 ~ "Unemployed", EMPLOY_D == 6 ~ "Not in labor force", TRUE ~ as.character(EMPLOY_D))) %>%
  group_by(EMPLOY_D) %>%
  count() %>%
  ggplot(aes(x = EMPLOY_D, y = n, fill = EMPLOY_D)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = EMPLOY_D), position = position_stack(vjust = 0.5), color = "black", size = 4) +
  coord_flip() +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(y = "# of Patients", x = "") + 
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, color = "black"), 
        axis.text.y = element_blank(), 
        axis.ticks.x = element_line(color = "black"), 
        axis.line.x = element_line(color = "black"), 
        panel.grid = element_blank(),
        legend.position = "none")


```
 
## Final: FREQ1_D Distribution
```{r}
data1 %>%
  group_by(FREQ1_D) %>%
  count() %>%
  ggplot(aes(x = FREQ1_D, y = n, fill = FREQ1_D)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = FREQ1_D), position = position_stack(vjust = 0.5), color = "black", size = 4) +
  coord_flip() +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(y = "# of Patients", x = "") + 
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, color = "black"), 
        axis.text.y = element_blank(), 
        axis.ticks.x = element_line(color = "black"), 
        axis.line.x = element_line(color = "black"), 
        panel.grid = element_blank(),
        legend.position = "none")

```
## Final LIVARAG_D Distribution
```{r}


# Load necessary libraries
library(ggplot2)
library(scales)

# Create a simplified bar chart
data1 %>%
  filter(LIVARAG_D != -9) %>%
  group_by(LIVARAG_D) %>%
  count() %>%
  ggplot(aes(x = LIVARAG_D, y = n, fill = LIVARAG_D)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = LIVARAG_D), position = position_stack(vjust = 0.5), color = "black", size = 4) +
  coord_flip() +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(y = "# of Patients", x = "") + 
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, color = "black"), 
        axis.text.y = element_blank(), 
        axis.ticks.x = element_line(color = "black"), 
        axis.line.x = element_line(color = "black"), 
        panel.grid = element_blank(),
        legend.position = "none")
```

