---
title: "FP5"
author: "Miles Libbey"
date: "2024-11-18"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(readr)
library(haven)
library(ggplot2)
library(ggalluvial)
library(ggmosaic)
library(purrr)
library(data.table)
library(tidymodels)
library(ranger)
library(vip)

set.seed(1111)
```

The purpose of this markdown is to explore the trends in housing status in state affiliated substance facilities.

## Loading Data

```{r echo=FALSE}
load("/Users/miles5/Desktop/tedsd_puf_2021_r.rdata")
tedsd_2021 <- tedsd_puf_2021_r
rm(tedsd_puf_2021_r)
```

## Setting up the Data Set

```{r, echo=FALSE}
political_grouping <- c(
    "1" = "Red", "2" = "Red", "5" = "Red", "8" = "Red", "9" = "Red", "17" = "Red", "18" = "Red",
    "20" = "Red", "21" = "Red", "22" = "Red", "28" = "Red", "29" = "Red", "30" = "Red", 
    "38" = "Red", "40" = "Red", "45" = "Red", "46" = "Red", "47" = "Red", "48" = "Red", 
    "49" = "Red", "54" = "Red", "56" = "Red",
    
    "6" = "Blue", "8" = "Blue", "9" = "Blue", "11" = "Blue", "15" = "Blue", "17" = "Blue", 
    "24" = "Blue", "25" = "Blue", "27" = "Blue", "34" = "Blue", "35" = "Blue", "36" = "Blue", 
    "50" = "Blue", "51" = "Blue", "53" = "Blue",
    
    "4" = "Swing", "12" = "Swing", "13" = "Swing", "26" = "Swing", "32" = "Swing", "33" = "Swing",
    "37" = "Swing", "39" = "Swing", "42" = "Swing", "55" = "Swing"
)

tedsd_2021 <- tedsd_2021 %>%
  mutate(Political_State = political_grouping[STFIPS]) %>%
  filter(
    LIVARAG != -9,
    LIVARAG_D != -9,
    REASON == 1,
    !is.na(Political_State)
  ) %>%
  mutate(across(everything(), as.factor)) %>%
    mutate(LIVARAG = recode(
    LIVARAG,
    `1` = "Unhoused",
    `2` = "Dependent Living",
    `3` = "Independent Living"
    )
  ) %>%
  mutate(LIVARAG_D = recode(
    LIVARAG_D,
    `1` = "Unhoused",
    `2` = "Dependent Living",
    `3` = "Independent Living"
    )) %>%
  mutate(LIVARAG = factor(LIVARAG, levels = c("Independent Living","Dependent Living","Unhoused"))) %>%
  mutate(LIVARAG_D = factor(LIVARAG_D, levels = c("Independent Living","Dependent Living","Unhoused")))
```

## Graph 1: Alluvial Plot Showing how Housing Status Changes in a Vacuum

```{r, fig.width=14}
tedsd_2021_alluvial <- tedsd_2021 %>%
    mutate(LIVARAG = recode(LIVARAG, #adding returns for readability / aesthetics
    "Unhoused" = "Unhoused",
    "Dependent Living" = "Dependent\n Living",
    "Independent Living" = "Independent\n Living",
    )
  ) %>%
  mutate(LIVARAG_D = recode(LIVARAG_D,
    "Unhoused" = "Unhoused",
    "Dependent Living" = "Dependent\n Living",
    "Independent Living" = "Independent\n Living",
    )
  ) %>%
  mutate(LIVARAG = factor(LIVARAG, levels = c("Independent\n Living", "Dependent\n Living", "Unhoused"))) %>%
  mutate(LIVARAG_D = factor(LIVARAG_D, levels = c("Independent\n Living", "Dependent\n Living", "Unhoused")))

tedsd_2021_alluvial %>%
  group_by(LIVARAG, LIVARAG_D) %>%
  summarize(count = n(), .groups = 'drop') %>%
  group_by(LIVARAG) %>%
  mutate(total = sum(count)) %>%
  ungroup() %>%
  mutate(percentage = count / total * 100) %>%
  ggplot(aes(axis1 = LIVARAG, axis2 = LIVARAG_D, y = percentage)) +
    geom_alluvium(aes(fill = LIVARAG)) +
    geom_stratum() +
    geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
    scale_x_discrete(limits = c("Admission", "Discharge"), expand = c(.12, .12)) +
    scale_fill_manual(values = c("#D55E00", "#56B4E9", "#009E73")) +
    labs(title = "Living Arrangements at Admission and Discharge by Substance",
         y = "Percentage",
         x = NULL) +
    theme_bw() + 
    theme(axis.text.x = element_text(size = 14), 
          axis.text.y = element_text(size = 14), 
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16), 
          plot.title = element_text(size = 18, face = "bold"),
          strip.text = element_text(size = 15),
          legend.position = "none"
    )
```

On its own, the most stable group is the independent living housing group, with a majority of population remaining the same 

## Graph 2: Alluvial Plot of Changes in Housing Status by Substance Used

```{r, fig.width=14}
overall <- tedsd_2021_alluvial %>%
  group_by(LIVARAG, LIVARAG_D) %>%
  summarize(count = n(), .groups = 'drop') %>%
  group_by(LIVARAG) %>%
  mutate(total = sum(count)) %>%
  ungroup() %>%
  mutate(percentage = count / total * 100) %>%
  mutate(SUB1_cat = "Overall")

stimulants <- tedsd_2021_alluvial %>%
  mutate(SUB1_cat = case_when(
    SUB1 == 2 ~ "Alcohol",
    SUB1 == 3 ~ "Stimulants",
    SUB1 == 5 ~ "Opiods",
    SUB1 == 7 ~ "Opiods",
    SUB1 == 10 ~ "Stimulants",
    SUB1 == 12 ~ "Stimulants",
    TRUE ~ NA_character_
    )
  ) %>% 
  filter(!is.na(SUB1_cat)) %>%
  group_by(SUB1_cat, LIVARAG, LIVARAG_D) %>%
  summarize(count = n(), .groups = 'drop') %>%
  group_by(SUB1_cat, LIVARAG) %>%
  mutate(total = sum(count)) %>%
  ungroup() %>%
  mutate(percentage = count / total * 100)
total_visual <- rbind(overall, stimulants)
total_visual%>%
  mutate(SUB1_cat = factor(SUB1_cat, levels = c("Alcohol", "Opiods", "Stimulants", "Overall"))) %>%
  ggplot(aes(axis1 = LIVARAG, axis2 = LIVARAG_D, y = percentage)) +
    geom_alluvium(aes(fill = LIVARAG)) +
    geom_stratum() +
    facet_wrap(~ SUB1_cat) +
    geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
    scale_x_discrete(limits = c("Admission", "Discharge"), expand = c(.12, .12)) +
    scale_fill_manual(values = c("#D55E00", "#56B4E9", "#009E73")) +
    labs(title = "Living Arrangements at Admission and Discharge by Substance",
         y = "Percentage",
         x = NULL) +
    theme_bw() + 
    theme(axis.text.x = element_text(size = 14), 
          axis.text.y = element_text(size = 14), 
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16), 
          plot.title = element_text(size = 18, face = "bold"),
          strip.text = element_text(size = 15),
          legend.position = "none"
    )
```

## Graph 3: Alluvial Plot Showing How Length of Stay Above 30 Days Impacts Changes in Housing

```{r, fig.width = 12, fig.height=8}
tedsd_2021 %>%
  # mutate(LOS = as.numeric(LOS)) %>% # Optional For Grouping LOS together
  #   mutate(LOS = cut(
  #   LOS,
  #   breaks = c(0, 30, 35, 37),
  #   labels = c("< 30 days", "30-180 days", "181 - 365+ days"),
  #   include.lowest = TRUE,
  #   right = TRUE
  # )) %>%
  mutate(LOS = case_when(
    LOS %in% 1:30 ~ "< 30 days",
    LOS == 31 ~ "31-45 days",
    LOS == 32 ~ "46-60 days",
    LOS == 33 ~ "61-90 days",
    LOS == 34 ~ "91-120 days",
    LOS == 35 ~ "121-180 days",
    LOS == 36 ~ "181-365 days",
    LOS == 37 ~ "365+ days"
    )
  ) %>%
  # mutate(LOS = factor(LOS, levels = c("< 30 days", "30-180 days", "181 - 365+ days"))) %>% #Optional for Grouping LOS together
  mutate(LOS = factor(LOS, levels = c("< 30 days", "31-45 days", "46-60 days", "61-90 days", "91-120 days", "121-180 days", "181-365 days", "365+ days"))) %>%
  group_by(LIVARAG, LOS, LIVARAG_D) %>%
  summarize(count = n(), .groups = 'drop') %>%
  group_by(LIVARAG, LOS) %>%
  mutate(total = sum(count)) %>%
  ungroup() %>%
  mutate(percentage = count / total * 100) %>% 
    mutate(LIVARAG_delta = case_when( #conditions on how housing changes from admission to discharge
    LIVARAG == "Unhoused" & LIVARAG_D == "Unhoused" ~ "Same",
    LIVARAG == "Unhoused" & LIVARAG_D == "Dependent Living" ~ "Went into Dependent Living",
    LIVARAG == "Unhoused" & LIVARAG_D == "Independent Living" ~ "Went into Independent Living",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Unhoused" ~ "Became Unhoused",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Dependent Living" ~ "Same",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Independent Living" ~ "Went into Independent Living",
    LIVARAG == "Independent Living" & LIVARAG_D == "Unhoused" ~ "Became Unhoused",
    LIVARAG == "Independent Living" & LIVARAG_D == "Dependent Living" ~ "Went into Dependent Living",
    LIVARAG == "Independent Living" & LIVARAG_D == "Independent Living" ~ "Same"
    )
  ) %>%
  mutate(LIVARAG_delta = factor(LIVARAG_delta, levels = c("Same", "Became Unhoused",
        "Went into Dependent Living", "Went into Independent Living"))
  ) %>%
  mutate(LIVARAG = recode(LIVARAG, #adding returns for readability / aesthetics
    "Unhoused" = "Lived Without Housing Before",
    "Dependent Living" = "Lived Dependently Before",
    "Independent Living" = "Came in Living Independently",
    )
  ) %>%
  ggplot(aes(x = LOS, y = percentage, fill = LIVARAG_delta)) +
    geom_bar(stat = "identity", position = "stack") +
    facet_wrap(~LIVARAG, ncol = 1, scales = "free_y") +
    scale_fill_manual(values = c("#0072B2", "#56B4E9", "#009E73", "#D55E00")) +
    labs(title = "Change in Clients' Housing Situation by Admitted Living Arrangement and Length of Stay",
         x = "Length of Stay (Above 30 days)",
         y = "Percentage of Clients (%)",
         fill = "Change in Housing Status") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
          axis.text.y = element_text(size = 12), 
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16), 
          plot.title = element_text(size = 16, face = "bold"),
          strip.text = element_text(size = 14),
          legend.title = element_text(size = 14),
          legend.text = element_text(size = 13))
```

## Graph 4: Faceted Scatter Plot Controlling for State Political Leanings on how Length of Stay impacts Changes in Being Unhoused

```{r}
tedsd_2021 %>%
  mutate(LIVARAG = factor(LIVARAG, levels = c("Independent Living", "Dependent Living", "Unhoused"))) %>%
  mutate(LIVARAG_delta = case_when(
    LIVARAG == "Unhoused" & LIVARAG_D == "Unhoused" ~ "Stays as Unhoused",
    LIVARAG == "Unhoused" & LIVARAG_D == "Dependent Living" ~ "to Dependent Living",
    LIVARAG == "Unhoused" & LIVARAG_D == "Independent Living" ~ "to Independent Living",
    TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(LIVARAG_delta)) %>% # just focusing on clients shown as unhoused when admitted
  mutate(
    LIVARAG_delta = factor(LIVARAG_delta, levels = c(
      "Stays as Unhoused",
      "to Dependent Living",
      "to Independent Living"
      )
    ),
    Political_State = factor(Political_State, levels = c("Blue", "Swing", "Red"))
  ) %>%
  group_by(LIVARAG_delta, Political_State, LOS) %>%
  summarize(count = n(), .groups = 'drop') %>% 
  group_by(Political_State, LOS) %>%
  mutate(proportion = count / sum(count)) %>%
  ggplot(aes(x = LOS, y = proportion, color = LIVARAG_delta, group = LIVARAG_delta)) +
    geom_smooth(se = FALSE, method = "loess", span = 0.15) + 
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = "Living Arrangements of Unhoused Individuals by State Political Color (2020 Electoral Map)",
      x = "Length of Stay (Days)",
      y = "Proportion",
      color = "Change from Unhoused"
    ) +
    facet_wrap(~Political_State, scales = "free_y") + 
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(size = 12),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10)
    )
```

# Mosaic 1: Changes in Living Arrangement by Length of Stay

```{r, fig.height=7,fig.width=7}
tedsd_2021 %>%
  mutate(LOS = as.numeric(LOS), 
         LOS = cut(LOS,
                   breaks = c(0, 30, 32, 37),
                   labels = c("Under 1 month", "30-90 days", "Over 90 days"),
                   include.lowest = TRUE,
                   right = TRUE
         )
  ) %>%
  mutate(LIVARAG_delta = case_when(
    LIVARAG == "Unhoused" & LIVARAG_D == "Unhoused" ~ "Same",
    LIVARAG == "Unhoused" & LIVARAG_D == "Dependent Living" ~ "Improved",
    LIVARAG == "Unhoused" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Dependent Living" ~ "Same",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Independent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Dependent Living" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Independent Living" ~ "Same"
  )) %>%
  count(LOS, LIVARAG_delta) %>%
  group_by(LOS) %>%
  mutate(proportion = n / sum(n)) %>%
  mutate(LIVARAG_delta = factor(LIVARAG_delta, levels = c("Worse", "Same", "Improved"))) %>%
  ggplot(aes(x = LOS, y = proportion, fill = LIVARAG_delta)) +
    geom_bar(stat = "identity", position = "fill") +
    geom_text(aes(label = scales::percent(proportion, accuracy = 0.1)),
              position = position_fill(vjust = 0.5), size = 5, fontface = "bold") +
    labs(title = "Impact of Length of Stay on Living Situation",
         subtitle = "Among Substance Use Patients in Public Facilities (2021)",
         x = "Length of Stay",
         y = "Percentage of Patients",
         fill = "Living Arrangement Change") +
    scale_fill_manual(values = c("Improved" = "darkgreen", "Same" = "gray", "Worse" = "red")) +
    theme_classic() +
    theme(legend.position = "bottom",
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
          plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11),
          axis.title.x = element_text(size = 14, face = "bold"),
          axis.text.x = element_text(size = 12, face = "bold"))
```

Interpretation: There seems to be a positive relationship between Length of Stay (LOS) and an upward change in housing status. On average, the longer a client stays in treatment, the more likely they are to move from being unhoused to living in independent housing (e.g., low-income or supportive housing) or from dependent housing to independently residing in an apartment or house of their own.
   When treatment durations are shorter, the proportions of clients experiencing upward and downward changes in housing status become more balanced. However, for clients with stays of more than 90 days, there is roughly the same number of individuals maintaining their current housing situation, but a greater proportion experience upward mobility, while significantly fewer see downward changes. This finding strongly supports the idea that longer treatment durations foster housing stability and improvement. Moreover, since this category includes lengths of stay extending well beyond 90 days (up to a year or more), it reinforces the trend of increased positive outcomes with longer treatment durations.


# Mosaic 2: Changes in Living Arrangement by Service received

```{r, fig.height=8,fig.width=6}
tedsd_2021 %>%
  mutate(SERVICES_grouped = case_when(
    SERVICES == 1 | SERVICES == 2 ~ "Detox",
    SERVICES == 3 | SERVICES == 4 | SERVICES == 5 ~ "Rehab/Residential",
    SERVICES == 6 | SERVICES == 7 | SERVICES == 8 ~ "Ambulatory"
  )) %>%
  mutate(LIVARAG_delta = case_when( 
    LIVARAG == "Unhoused" & LIVARAG_D == "Unhoused" ~ "Same",
    LIVARAG == "Unhoused" & LIVARAG_D == "Dependent Living" ~ "Improved",
    LIVARAG == "Unhoused" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Dependent Living" ~ "Same",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Independent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Dependent Living" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Independent Living" ~ "Same"
  )) %>%
  count(SERVICES_grouped, LIVARAG_delta) %>%
  group_by(SERVICES_grouped) %>%
  mutate(proportion = n / sum(n)) %>%
  mutate(LIVARAG_delta = factor(LIVARAG_delta, levels = c("Worse", "Same", "Improved"))) %>%
  ggplot(aes(x = SERVICES_grouped, y = proportion, fill = LIVARAG_delta)) +
    geom_bar(stat = "identity", position = "fill") +
    geom_text(aes(label = scales::percent(proportion, accuracy = 0.1)),
              position = position_fill(vjust = 0.5), size = 5, fontface = "bold") +
    labs(title = "Impact of Treatment Service on Living Situation",
         subtitle = "Among Substance Use Patients in Public Facilities (2021)",
         x = "Treatment Service",
         y = "Percentage of Patients",
         fill = "Living Arrangement Change") +
    scale_fill_manual(values = c("Improved" = "darkgreen", "Same" = "gray", "Worse" = "red")) +
    theme_classic() +
    theme(legend.position = "bottom",
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
          plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11),
          axis.title.x = element_text(size = 14, face = "bold"),
          axis.text.x = element_text(size = 12, face = "bold"))
```
Interpretation: Clients participating in detoxification or rehab/residential services appear more likely to experience a positive shift in their housing compared to those receiving detoxification services. This trend can be attributed to two main factors. First, as suggested by the Length of Stay Mosaic, the longer a client remains in treatment, the more likely they are to experience upward mobility in their living arrangements. Rehabilitation and residential services often extend for much longer durations, allowing more time for improvements in housing stability. Second, there is literature that suggests that detoxification and rehab/residential services attract individuals who are unhoused or living in supportive housing. If someone is already living independently, there is less opportunity for upward housing mobility, meaning the observed improvements are concentrated among those starting in less stable housing situations.This does not, however, diminish the significant positive impact of ambulatory and rehab/residential services. These services are strongly correlated with upward shifts in housing and greatly outweigh any negative outcomes.

# SERVICES : LOS Mosaic

```{r}
# NOT AN ACTUAL GRAPH; Just Context for other Graphs
tedsd_2021 %>%
  mutate(
    LOS = as.numeric(LOS),
    LOS = cut(
      LOS,
      breaks = c(0, 30, 32, 37),
      labels = c("Under 1 month", "30-90 days", "Over 90 days"),
      include.lowest = TRUE,
      right = TRUE
    )
  ) %>%
  mutate(SERVICES_grouped = case_when(
    SERVICES == 1 | SERVICES == 2 ~ "Detox, 24 hour",
    SERVICES == 3 | SERVICES == 4 | SERVICES == 5 ~ "Rehab/Residential",
    SERVICES == 6 | SERVICES == 7 | SERVICES == 8 ~ "Ambulatory"
  )) %>%
  mutate(LIVARAG_delta = case_when(
    LIVARAG == "Unhoused" & LIVARAG_D == "Unhoused" ~ "Same",
    LIVARAG == "Unhoused" & LIVARAG_D == "Dependent Living" ~ "Improved",
    LIVARAG == "Unhoused" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Dependent Living" ~ "Same",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Independent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Dependent Living" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Independent Living" ~ "Same"
  )) %>%
  mutate(LIVARAG_delta = factor(LIVARAG_delta, levels = c("Improved", "Same", "Worse"))) %>%
  mutate(LOS = factor(LOS, levels = c("Under 1 month", "30-90 days", "Over 90 days"))) %>%
  mutate(SERVICES_LOS = paste(SERVICES_grouped, LOS, sep = ": ")) %>%
  ggplot() + # Credit to Avianna
    geom_mosaic(aes(x = product(LIVARAG_delta, SERVICES_LOS), fill = LIVARAG_delta)) +
    geom_mosaic_text(aes(x = product(LIVARAG_delta, SERVICES_LOS), label = scales::percent(after_stat(.wt)/ sum(after_stat(.wt))))) +
     labs(fill = "Living Situation Change", 
          y = "Proportion of Substance Use Patients", 
          x = "Type of Treatment Received", 
          title = "Impact of Housing Situation on Type of Treatment Outcome", 
          subtitle = "Among Substance Use Patients in Public Facilities (2021)") +
    #scale_fill_manual(values = c("#2A788E", "#FDE725")) +
    scale_fill_manual(
      values = c(
        "Improved" = "darkgreen",
        "Same" = "gray",
        "Worse" = "red")
    ) +
    theme_classic() +
    coord_flip() +
    theme(legend.position = "bottom", 
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
          plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11))
```

# Proportional LOS Bar Plot

```{r}
tedsd_2021 %>%
  mutate(
    LOS = as.numeric(LOS),
    LOS = cut(
      LOS,
      breaks = c(0, 30, 32, 37),
      labels = c("Under 1 month", "30-90 days", "Over 90 days"),
      include.lowest = TRUE,
      right = TRUE
    )
  ) %>%
  mutate(LIVARAG_delta = case_when( 
    LIVARAG == "Unhoused" & LIVARAG_D == "Unhoused" ~ "Same",
    LIVARAG == "Unhoused" & LIVARAG_D == "Dependent Living" ~ "Improved",
    LIVARAG == "Unhoused" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Dependent Living" ~ "Same",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Independent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Dependent Living" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Independent Living" ~ "Same"
  )) %>%
  mutate(LIVARAG_delta = factor(LIVARAG_delta, levels = c("Improved", "Same", "Worse"))) %>%
  group_by(LOS, LIVARAG_delta) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(LOS) %>%
  mutate(
    proportion = count / sum(count),
    label = scales::percent(proportion, accuracy = 0.1)
  ) %>%
  ggplot(aes(x = LOS, y = proportion, fill = LIVARAG_delta)) +
    geom_bar(stat = "identity", position = "fill") +
    geom_text(
      aes(label = label),
      position = position_fill(vjust = 0.5),
      size = 3
    ) +
    labs(
      title = "Impact of Housing Situation on Length of Stay Outcome",
      x = "Length of Treatment",
      y = "Proportion of Patients",
      fill = "Living Arrangement Change"
    ) +
    scale_fill_manual(
      values = c(
        "Improved" = "darkgreen",
        "Same" = "gray",
        "Worse" = "red")
    ) +
    theme_classic() +
    coord_flip() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11)
  )
```

# Proportional SERVICES Bar Plot

```{r}
tedsd_2021 %>%
  mutate(SERVICES_grouped = case_when(
    SERVICES == 1 | SERVICES == 2 ~ "Detox, 24 hour",
    SERVICES == 3 | SERVICES == 4 | SERVICES == 5 ~ "Rehab/Residential",
    SERVICES == 6 | SERVICES == 7 | SERVICES == 8 ~ "Ambulatory"
  )) %>%
  mutate(LIVARAG_delta = case_when( 
    LIVARAG == "Unhoused" & LIVARAG_D == "Unhoused" ~ "Same",
    LIVARAG == "Unhoused" & LIVARAG_D == "Dependent Living" ~ "Improved",
    LIVARAG == "Unhoused" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Dependent Living" ~ "Same",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Independent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Dependent Living" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Independent Living" ~ "Same"
  )) %>%
  mutate(LIVARAG_delta = factor(LIVARAG_delta, levels = c("Worse", "Same", "Improved"))) %>%
  mutate(SERVICES_grouped = factor(SERVICES_grouped, levels = c("Detox, 24 hour", "Rehab/Residential", "Ambulatory"))) %>%
  group_by(SERVICES_grouped, LIVARAG_delta) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(SERVICES_grouped) %>%
  mutate(
    proportion = count / sum(count),
    label = scales::percent(proportion, accuracy = 0.1)
  ) %>%
  ggplot(aes(x = SERVICES_grouped, y = proportion, fill = LIVARAG_delta)) +
    geom_bar(stat = "identity", position = "fill") +
    geom_text(
      aes(label = label),
      position = position_fill(vjust = 0.5),
      size = 3
    ) +
    labs(
      title = "Impact of Treatment Services on Living Situation Outcome",
      x = "Length of Treatment",
      y = "Proportion of Patients",
      fill = "Living Arrangement Change"
    ) +
    scale_fill_manual(
      values = c(
        "Improved" = "darkgreen",
        "Same" = "gray",
        "Worse" = "red")
    ) +
    guides(fill = guide_legend(reverse = TRUE)) + 
    theme_classic() +
    coord_flip() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11)
  )
```

# Proportional SERVICES : LOS Bar Plot

```{r}
tedsd_2021 %>%
  mutate(
    LOS = as.numeric(LOS),
    LOS = cut(
      LOS,
      breaks = c(0, 30, 32, 37),
      labels = c("Under 1 month", "30-90 days", "Over 90 days"),
      include.lowest = TRUE,
      right = TRUE
    )
  ) %>%
  mutate(SERVICES_grouped = case_when(
    SERVICES == 1 | SERVICES == 2 ~ "Detox, 24 hour",
    SERVICES == 3 | SERVICES == 4 | SERVICES == 5 ~ "Rehab/Residential",
    SERVICES == 6 | SERVICES == 7 | SERVICES == 8 ~ "Ambulatory"
  )) %>%
  mutate(LIVARAG_delta = case_when(
    LIVARAG == "Unhoused" & LIVARAG_D == "Unhoused" ~ "Same",
    LIVARAG == "Unhoused" & LIVARAG_D == "Dependent Living" ~ "Improved",
    LIVARAG == "Unhoused" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Dependent Living" ~ "Same",
    LIVARAG == "Dependent Living" & LIVARAG_D == "Independent Living" ~ "Improved",
    LIVARAG == "Independent Living" & LIVARAG_D == "Unhoused" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Dependent Living" ~ "Worse",
    LIVARAG == "Independent Living" & LIVARAG_D == "Independent Living" ~ "Same"
  )) %>%
  mutate(LIVARAG_delta = factor(LIVARAG_delta, levels = c("Improved", "Same", "Worse"))) %>%
  mutate(SERVICES_LOS = paste(SERVICES_grouped, LOS, sep = ": ")) %>%
  group_by(SERVICES_LOS, LIVARAG_delta) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(SERVICES_LOS) %>%
  mutate(
    proportion = count / sum(count),
    label = scales::percent(proportion, accuracy = 0.1)
  ) %>%
  ggplot(aes(x = SERVICES_LOS, y = proportion, fill = LIVARAG_delta)) +
    geom_bar(stat = "identity", position = "fill") +
    geom_text(
      aes(label = label),
      position = position_fill(vjust = 0.5),
      size = 3
    ) +
    labs(
      title = "Impact of Housing Situation on Services and Length of Stay Outcome",
      x = "Services and Length of Stay",
      y = "Proportion of Patients",
      fill = "Living Arrangement Change"
    ) +
    scale_fill_manual(
      values = c(
        "Improved" = "darkgreen",
        "Same" = "gray",
        "Worse" = "red")
    ) +
    theme_classic() +
    coord_flip() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.subtitle = element_text(face = "italic", hjust = 0.5, size = 11)
    )
```
```{r}
tedsd_2021 %>%
  cor(use = "complete.obs") %>%          
  as.data.frame() %>%                              
  rownames_to_column("Variable") %>%        
  filter(Variable == "TOPO") %>%             
  pivot_longer(-Variable, names_to = "Other_Variable", values_to = "Correlation") %>% 
  arrange(desc(Correlation))
```

# Random Forest Classification

```{r, fig.height=6, fig.width=4}
modelDataSubset <- tedsd_2021 %>%
  mutate(SUB2 = ifelse(SUB2 == 1, 1, 0)) %>%
  mutate(SUB3 = ifelse(SUB3 == 1, 1, 0)) %>%
  mutate(SERVICES = case_when(
  SERVICES == 1 | SERVICES == 2 ~ "Detox, 24 hour",
  SERVICES == 3 | SERVICES == 4 | SERVICES == 5 ~ "Rehab/Residential",
  SERVICES == 6 | SERVICES == 7 | SERVICES == 8 ~ "Ambulatory"
  )) %>%
  select(LIVARAG_D, AGE, GENDER, ETHNIC, EDUC, RACE, MARSTAT, VET, LIVARAG, EMPLOY, EMPLOY_D, FREQ1, FREQ1_D, FRSTUSE1, DSMCRIT, PSYPROB, PSOURCE, STFIPS, HLTHINS, ARRESTS_D, NOPRIOR, SUB1, SUB2, SUB3, METHUSE, LOS, SERVICES, REASON, FREQ_ATND_SELF_HELP_D)

rf_spec <- rand_forest() %>%
  set_engine(engine = 'ranger') %>% 
  set_args(mtry = NULL, 
           trees = 500, 
           min_n = 2,
           probability = FALSE, 
           importance = 'impurity') %>% 
  set_mode('classification') 

data_rec <- recipe(LIVARAG_D ~ ., data = modelDataSubset) # Predicting Living arrangment at Discharge

data_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(data_rec)

rf_fit <- fit(data_wf, data = modelDataSubset)

rf_fit

#Variable importance Plot
rf_fit %>%
    extract_fit_engine() %>%
    vip(num_features = 6, mapping = aes(fill = .data[["Variable"]])) +
    scale_fill_manual(values = c("grey", "grey", "grey" ,"#A94064" ,"#A94064" ,"grey" )) +
    labs(title = "Top 6 Most Important Variables ",
         subtitle = "in Predicting Living Situation at Discharge") +
    theme_classic() +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", size = 11),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(), 
          axis.line.y = element_blank()) +
    geom_text(aes(label = c("HOUSING AT ADMISSION",
                            "LENGTH OF STAY",
                              "AGE",
                              "PSYCH DIAGNOSES",
                              "STATE",
                            "SERVICES AT ADMISSION"),
                            y = 0), hjust = 0, color = "black", size = 4, fontface = "bold")

#accuracy of random forest model
rf_OOB_output <- function(fit_model, model_label, truth){
    tibble(
          .pred_Unemployed = fit_model %>% extract_fit_engine() %>% pluck('predictions'), #OOB predictions
          LIVARAG_D = truth,
          label = model_label
      )
}

output <- rf_OOB_output(rf_fit, "test", modelDataSubset  %>% pull(LIVARAG_D))
output %>% 
    accuracy(truth = LIVARAG_D, estimate = .pred_Unemployed)
```

# Hyper-Parameter Tuning

```{r}
rf_spec <- rand_forest() %>%
  set_engine(engine = 'ranger') %>% 
  set_mode('classification')

data_rec <- recipe(LIVARAG_D ~ ., data = modelDataSubset)

data_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(data_rec)

rf_grid <- grid_random( #Grid to tune to
  trees(c(300,1500)),
  min_n(c(2, 7)) 
)

cv_folds <- vfold_cv(modelDataSubset, v = 5)


tune_results <- tune_grid( # Actually tuning the model
  object = data_wf,
  resamples = cv_folds, 
  grid = rf_grid, 
  metrics = metric_set(accuracy)
)

tune_results

best_rf <- select_best(tune_results, metric = "accuracy")

final_rf <- finalize_workflow(data_wf, best_rf)

final_rf_fit <- fit(final_rf, data = modelDataSubset)
final_rf_fit
```

